\normalfont \normalsize \textsc{University of Stuttgart} \\ [25pt] \horrule{0.5pt} \\[0.4cm] \huge Hauptseminar:\\The Logjam Attack \\ \horrule{2pt} \\[0.5cm]

### 2017-06-12 Ausarbeitung/paper.tex
precomputation for NFS improved

\subsection{Precomputation}

\subsubsection{Smoothness Assumption}
Define that an integer number $x$ is $B$-smooth, if all prime factors of $x$ are smaller that $B$. NFS assumes the base $\alpha$ to be $B$-smooth where $B$ is a upper bound for the prime values in the factor base (which is the output of the precomputation). If $\alpha$ is not $B$-smooth, we can use a $B$-smooth $\alpha'$ instead. Then, we can compute $\log_{\alpha'}\alpha$ and transform back to the original problem with
\begin{equation}
\log_\alpha D\equiv\frac{\log_{\alpha'}D}{\log_{\alpha'}\alpha}\mod (q-1)
\text{ .}
\end{equation}
If $\alpha'$ is not a generator for $\mathrm{GF}(q)^*$, the logs on the right hand side might not exist. $\alpha'$ can be checked for feasibility by factoring $(q-1)$ with the NFS factoring algorithm and check
\begin{equation}
(a')^{\frac{q-1}{p}} \not\equiv 1 \mod q,\quad\forall p\parallel q
\text{ ,}
\end{equation}
where $p\parallel q$ denotes that $p$ is a prime divisor of $q$. The reason why $\alpha'$ needs to be $B$-Smooth is that the factor base needs to have at least one inhomogeneous relation for the logs of the factor base, using the equation
\begin{eqnarray}
\label{eqn:assumption}
\log_\alpha\alpha = 1 \equiv \sum_{p^t\parallel a}t\log_\alpha p \mod (q-1)
\text{ ,}
\end{eqnarray}
otherwise we gain no knowledge from the factor base.

\subsubsection{Polynomial Selection}
The goal in this step to find a polynomial representation of $\mathrm{GF}(q)$. This has the advantage that we can express relations and encode informations in terms of linear equations to reason about them, which are the tasks that actually will be done in the next steps.

@@

\subsubsection{Sieving}
In this step we try to find linear equations that are informative in a sense that they are related to the the logarithms in $\mathcal{B}$. For that, sieve through integer pairs $(c,d)$ that are co-prime for which $c+dm$ and $c+d\gamma$ are $S$-smooth where $S = \prod_{s\in\mathcal{B}_\mathbb{Q}}$. For this smoothness criteria, we can formulate equations
\begin{eqnarray}
c+dm = \prod_{s\in\mathcal{B}_\mathbb{Q}}s^{w_s(c,d)}
\end{eqnarray}
and
\begin{eqnarray}
|N(c+d\gamma)| = \prod_{s\in\mathcal{B}_\mathbb{Q}}s^{v_s(c,d)}
\text{ ,}
\end{eqnarray}
where $N(c+d\gamma) = (-d)^kf(-\frac{c}{d})$ (see original paper \citep{gordon1993discrete} for explanation), for some $w_s,v_s\in\mathbb{Z}^+$.

From here we can derive factorizations for primes in $\mathcal{B}_K$ with the fact (not proven here) that there is an unique ideal $\hat{s}\in\mathcal{B}_K$ lying over $s$ and dividing $c+d\gamma$. Let $v_{\hat{S}}(c,d) = v_s(c,d)$ for this ideal and be zero for other ideals in $\mathcal{B}_K$ of norm $s$. Then we have the following
\begin{eqnarray}
\label{eqnSieve1}
c+dm = \prod_{s\in\mathcal{B}_\mathbb{Q}}s^{w_s(c,d)}
\end{eqnarray}
and
\begin{eqnarray}
\label{eqnSieve2}
c+d\gamma = \prod_{\hat{s}\in\mathcal{B}_K}\hat{s}^{v_{\hat{s}}(c,d)}
\text{ .}
\end{eqnarray}
We repeats the steps described so far, which means, try to find more pairs $(c,d)$ to build equations \ref{eqnSieve1} and \ref{eqnSieve2} until we have at least $|\mathcal{B}|$ of these equations. These are now used to form a large matrix $A$, where its rows are the $w_s$'s and $v_{\hat{s}}$'s for each equation. From $A$ we extract only those columns where elements are in $\mathcal{B}-\mathcal{B}'$, forming a new matrix $A'$. This means, we are only interested in solutions for primes $s\in\mathcal{B}'$. $A'$ is now the input for the next stage.

\subsubsection{Linear Algebra}
Given some matrix $A = \mathbb{Z}^{S\times T},\ S > T$, where each row has at most $E$ non-zero entries, find a linear relation over $\mathbb{Q}$ for the rows of $A$. This is accomplished in four steps, shown for the case $S=T+1$. This procedure was developed by Carl Pomerance (*1944) which was not separately published, instead only for helping out in \citep{gordon1993discrete}.
\paragraph{Step 1}Attempt to compute the rank $r = \mathrm{rk}(A)$. For that, choose a random prime $q_0\leq ET\log T$. Later, we will check if that choice was feasible. If not, come back here and sample another one. Then, use Gaussian elimination mod $q_0$ to find the rank $r_0$ of $A\mod q_0$. Rearrange the rows so that the first $r_0$ rows are linearly independent mod $q_0$. Now the rearranged matrix is
\begin{equation}
A' =
\begin{pmatrix}
v_1\\v_2\\\vdots\\v_{T\!+\!1}
\end{pmatrix}
=
\begin{pmatrix}
a'_{11} & a'_{12} & \cdots & a'_{1T} \\
a'_{21} & a'_{22} & \cdots & a'_{2T} \\
\vdots & \vdots & \ddots & \vdots \\
a'_{T\!+\!11} & a'_{T\!+\!12} & \cdots & a'_{T\!+\!1T} \\
\end{pmatrix}
\text{ .}
\end{equation}
The result of the Gaussian elimination provide a submatrix $\hat{A} = \mathbb{Z}^{r_0\times r_0}$ of the first $r_0$ rows that is nonsingular mod $q_0$,
\begin{equation}
\hat{A} =
\begin{pmatrix}
v_1\\v_2\\\vdots\\v_{r_0}
\end{pmatrix}
=
\begin{pmatrix}
a'_{11} & a'_{12} & \cdots & a'_{1r_0} \\
a'_{21} & a'_{22} & \cdots & a'_{2r_0} \\
\vdots & \vdots & \ddots & \vdots \\
a'_{r_01} & a'_{r_02} & \cdots & a'_{r_0r_0} \\
\end{pmatrix}
\text{ .}
\end{equation}
\paragraph{Step 2} Attempt to express $v_{r_0+1}$ as a linear combination of $v_1,v_2,...,v_{r_0}\mod q$ for each $q\leq ET\log T$. For some $q$ it will be successful, for others not. Denote the product of all successful primes as $\boldsymbol{P}$ and the product of all unsuccessful primes as $\boldsymbol{P'}$. If $\boldsymbol{P'} > ET\log T$, then repeat Step 1 (sample a new random prime number). Otherwise, continue with the next step.
\paragraph{Step 3} Attempt to find the determinant $D$ of $\hat{A}$. For each prime $q|\boldsymbol{P}$, use the Wieldemann's probabilistic determinant algorithm \citep{wiedemann1986solving} to compute an integer $D_q\in\{0,...,q-1\}$, which is the determinant of $A\mod q$ with probability of at least $1-(ET)^{-2}$, i.e.
\begin{equation}
\mathbb{P}(D_q = \det(A\mod q)\mid A,q) \geq 1-(ET)^{-2}
\text{ .}
\end{equation}
For each $D_q$ we can write an congruence
\begin{equation}
D_0 \equiv D_q \mod q,\quad\forall q|\boldsymbol{P}
\end{equation}
and use the Chinese remainder theorem to compute $D_0$ fulfilling these congruences and being closest to zero. Repeat this step until we have found a value that lies within the bound
\begin{equation}
0 < |D_0| \leq (E^{\frac{1}{2}}T)^T
\text{ .}
\end{equation}

@@

After applying the algorithm described in the Linear Algebra step on the matrix resulted form the Sieving step, we have some new equations
\begin{equation}
\prod_{(c,d)}(c+dm)^{x(c,d)} = U
\textit{ ,}
\end{equation}
where $x(c,d)$ are some newly computed integers and $U$ is a unit in $O_K$. After some intermediate steps, we can express the logs of primes $s\in\mathcal{B}$ as $z_s$ (which is here a black box) and state
\begin{equation}
\prod_{s\in\mathcal{B}'}s^{z_s}\equiv 1\mod q
\text{ .}
\end{equation}
Taking the log, we have
\begin{equation}
\sum_{s\in\mathcal{B}'}\log_\alpha s \equiv 0 \mod (q-1)
\text{ .}
\end{equation}
With more than $|\mathcal{B}'|$ such congruences, we can solve these with equation \ref{eqn:assumption} and obtain the logs for all primes $s\in\mathcal{B}'$.

>>>>

The algorithm for NFS in general is technically difficult to understand. That means there are many mathematical objects we must deal with and keep in mind at the same time. For simplicity I neglected some details for a better reading, while the drawback is that I need to refer to the original paper from time to time.

\subsection{Precomputation}
The precomputation consists of three steps. In Polynomial Selection, we find an appropriate representation for $\mathrm{GF}(q)$. Using this representation, we find equations that relate certain numbers with smaller primes in $\mathrm{GF}(q)$ in the Sieving step. These equations form matrices that can be used to find relations that lead to the logs of the base $\alpha$ in the Linear Algebra step. But first, one assumption must be made on $\alpha$.

\subsubsection{Smoothness Assumption}
Define that an integer number $x$ is $B$-smooth, if all prime factors of $x$ are smaller that $B$. NFS assumes the base $\alpha$ to be $B$-smooth where $B$ is an upper bound for the prime values in the factor base (which is the output of the precomputation). If $\alpha$ is not $B$-smooth, we can use a $B$-smooth $\alpha'$ instead. Then, we can compute $\log_{\alpha'}\alpha$ and transform back to the original problem with
\begin{equation}
\log_\alpha D\equiv\frac{\log_{\alpha'}D}{\log_{\alpha'}\alpha}\mod (q-1)
\text{ .}
\end{equation}
If $\alpha'$ is not a generator for $\mathrm{GF}(q)^*$, the logs on the right hand side might not exist. $\alpha'$ can be checked for feasibility by factoring $(q-1)$ with the NFS factoring algorithm and check
\begin{equation}
(a')^{\frac{q-1}{p}} \not\equiv 1 \mod q,\quad\forall p\parallel q
\text{ ,}
\end{equation}
where $p\parallel q$ denotes that $p$ is a prime divisor of $q$. The reason why $\alpha'$ needs to be $B$-Smooth is that the factor base needs to have at least one inhomogeneous relation for the logs of the factor base, using the equation
\begin{eqnarray}
\label{eqn:assumption}
\log_\alpha\alpha = 1 \equiv \sum_{p^t\parallel a}t\log_\alpha p \mod (q-1)
\text{ ,}
\end{eqnarray}
otherwise we gain no knowledge from the factor base.

\subsubsection{Polynomial Selection}
The goal in the first step is to find a polynomial representation of $\mathrm{GF}(q)$. This has the advantage that we can express relations and encode informations in terms of linear equations to reason about them, which are the tasks that actually will be done in the next steps.

@@

\subsubsection{Sieving}
In this step we try to find linear equations that are informative in the sense that they are related to the the logarithms in $\mathcal{B}$. For that, sieve through integer pairs $(c,d)$ that are co-prime for which $c+dm$ and $c+d\gamma$ are $S$-smooth where $S = \prod_{s\in\mathcal{B}_\mathbb{Q}}$. For this smoothness criteria, we can formulate equations
\begin{eqnarray}
c+dm = \prod_{s\in\mathcal{B}_\mathbb{Q}}s^{w_s(c,d)}
\end{eqnarray}
and
\begin{eqnarray}
|N(c+d\gamma)| = \prod_{s\in\mathcal{B}_\mathbb{Q}}s^{v_s(c,d)}
\text{ ,}
\end{eqnarray}
where $N(c+d\gamma) = (-d)^kf(-\frac{c}{d})$ (see original paper \citep{gordon1993discrete} for explanation), for some $w_s,v_s\in\mathbb{Z}^+$.

From here we can derive factorizations for primes in $\mathcal{B}_K$ with the fact (not proven here) that there is an unique ideal $\hat{s}\in\mathcal{B}_K$ lying over $s$ and dividing $c+d\gamma$. Let $v_{\hat{S}}(c,d) = v_s(c,d)$ for this ideal and be zero for other ideals in $\mathcal{B}_K$ of norm $s$. Then we have the following
\begin{eqnarray}
\label{eqnSieve1}
c+dm = \prod_{s\in\mathcal{B}_\mathbb{Q}}s^{w_s(c,d)}
\end{eqnarray}
and
\begin{eqnarray}
\label{eqnSieve2}
c+d\gamma = \prod_{\hat{s}\in\mathcal{B}_K}\hat{s}^{v_{\hat{s}}(c,d)}
\text{ .}
\end{eqnarray}
Repeats the steps described so far to find more pairs $(c,d)$ to build equations \ref{eqnSieve1} and \ref{eqnSieve2} until we have at least $|\mathcal{B}|$ of these equations. For each equation, build a matrix, where its rows are the $w_s$'s and $v_{\hat{s}}$'s, we call this set $\mathcal{A}$. We modify $\mathcal{A}$ by extracting only those columns where elements are in $\mathcal{B}-\mathcal{B}'$. This means, we are only interested in solutions for primes $s\in\mathcal{B}'$. $\mathcal{A}$ is now the input for the next step.

\subsubsection{Linear Algebra}
For each $A \in \mathcal{A},\ A = \mathbb{Z}^{S\times T},\ S > T$, where each row has at most $E$ non-zero entries, find a linear relation over $\mathbb{Q}$ for the rows of $A$. This is accomplished in four steps, shown for the case $S=T+1$. This procedure was developed by Carl Pomerance (*1944) which was not separately published, instead only for helping out in \citep{gordon1993discrete}.
\paragraph{Step 1}Attempt to compute the rank $r = \mathrm{rk}(A)$. For that, choose a random prime $q_0\leq ET\log T$. Later, we will check if that choice was feasible. If not, come back here and sample another one. Then, use Gaussian elimination mod $q_0$ to find the rank $r_0$ of $A\mod q_0$. Rearrange the rows so that the first $r_0$ rows are linearly independent mod $q_0$. Now the rearranged matrix is
\begin{equation}
A' =
\begin{pmatrix}
v_1\\v_2\\\vdots\\v_{T\!+\!1}
\end{pmatrix}
=
\begin{pmatrix}
a'_{11} & a'_{12} & \cdots & a'_{1T} \\
a'_{21} & a'_{22} & \cdots & a'_{2T} \\
\vdots & \vdots & \ddots & \vdots \\
a'_{T\!+\!11} & a'_{T\!+\!12} & \cdots & a'_{T\!+\!1T} \\
\end{pmatrix}
\text{ .}
\end{equation}
The result of the Gaussian elimination provide a submatrix $\hat{A} = \mathbb{Z}^{r_0\times r_0}$ of the first $r_0$ rows that is nonsingular mod $q_0$,
\begin{equation}
\hat{A} =
\begin{pmatrix}
v_1\\v_2\\\vdots\\v_{r_0}
\end{pmatrix}
=
\begin{pmatrix}
a'_{11} & a'_{12} & \cdots & a'_{1r_0} \\
a'_{21} & a'_{22} & \cdots & a'_{2r_0} \\
\vdots & \vdots & \ddots & \vdots \\
a'_{r_01} & a'_{r_02} & \cdots & a'_{r_0r_0} \\
\end{pmatrix}
\text{ .}
\end{equation}
\paragraph{Step 2} Attempt to express $v_{r_0+1}$ as a linear combination of $v_1,v_2,...,v_{r_0}\mod q$ for each $q\leq ET\log T$. For some $q$ it will be successful, for others not. Denote the product of all successful primes as $\boldsymbol{P}$ and the product of all unsuccessful primes as $\boldsymbol{P'}$. If $\boldsymbol{P'} > ET\log T$, then repeat Step 1 (sample a new random prime number). Otherwise, continue with the next step.
\paragraph{Step 3} Attempt to find the determinant $D$ of $\hat{A}$. For each prime $q|\boldsymbol{P}$, use the Wieldemann's probabilistic determinant algorithm \citep{wiedemann1986solving} to compute an integer $D_q\in\{0,...,q-1\}$, which is the determinant of $A\mod q$ with probability of at least $1-(ET)^{-2}$, i.e.
\begin{equation}
\mathbb{P}(D_q = \det(A\mod q)\mid A,q) \geq 1-(ET)^{-2}
\text{ .}
\end{equation}
For each $D_q$ we can write an congruence
\begin{equation}
D_0 \equiv D_q \mod q,\quad\forall q|\boldsymbol{P}
\end{equation}
and use the Chinese remainder theorem to compute $D_0$ fulfilling these congruences and being closest to zero. Repeat this step until we have found a value that lies within the bound
\begin{equation}
0 < |D_0| \leq (E^{\frac{1}{2}}T)^T
\text{ .}
\end{equation}

@@

After applying the algorithm described in the Linear Algebra step on the matrices resulted form the Sieving step, we have some new equations
\begin{equation}
\prod_{(c,d)}(c+dm)^{x(c,d)} = U
\textit{ ,}
\end{equation}
where $x(c,d)$ is the solution for the matrix constructed from $(c,d)$ and $U$ is a unit in $O_K$. After some intermediate steps, we can express the logs of primes $s\in\mathcal{B}$ as $z_s$ (which is a black box here) and state
\begin{equation}
\prod_{s\in\mathcal{B}'}s^{z_s}\equiv 1\mod q
\text{ .}
\end{equation}
Taking the log, we have
\begin{equation}
\sum_{s\in\mathcal{B}'}z_s\log_\alpha s \equiv 0 \mod (q-1)
\text{ .}
\end{equation}
With more than $|\mathcal{B}'|$ such congruences, we can solve these with equation \ref{eqn:assumption} and obtain the logs for all primes $s\in\mathcal{B}'$.
